---
title: "Advanced Multivariate Dynamics of Canadian Mortality (2000–2024)"
subtitle: "An Integrated Feature-Based Clustering, Penalized Regression, and Hierarchical Forecasting Pipeline"
author: "Margaret Mangwende, Hanping Wang and Definate Mudarikwa"
format: 
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show code"
    fig-width: 9
    fig-height: 6
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---

# 1. Introduction and Background

Canada is currently undergoing a profound demographic shift. With the rapid aging of the baby boomer generation, shifting post-pandemic health baselines, and increasingly strained healthcare infrastructure, understanding the mechanics of national mortality was not treated as just an academic exercise. Statistics Canada data reveals that while life expectancy generally continues to climb, the sheer volume of mortality is increasing, driven largely by the 85-and-older cohort. However, this overarching trend hides highly complex, diverging trajectories beneath the surface depending on a patient's biological sex, precise age group, and underlying conditions.

## 1.1 Problem Statement

Historically, public health reporting has relied heavily on static dashboards and simple year-over-year counts. While these methods show what happened, they fail to capture the multidimensional behavior of diseases over time. They cannot mathematically distinguish between a chronic, compounding threat (like Alzheimer's) and a highly volatile, sudden-onset event (like COVID-19 or respiratory spikes). Furthermore, raw counts obscure the true underlying risk multipliers of specific demographic traits. Without isolating these hidden mathematical features and projecting them forward, healthcare resource allocation remains reactive rather than proactive.

## 1.2 Main Aim

The primary aim of this analysis is to transition Canadian mortality data from a descriptive reporting framework into an advanced, predictive machine learning pipeline. By treating causes of death as multidimensional time-series problems, we aim to uncover structural behavioral patterns, isolate true demographic risk multipliers, and generate robust forecasts for the coming decade.

## 1.3 Objectives and Research Questions

To achieve this aim, the analysis is structured around four core objectives and corresponding research questions:

Objective 1: Establish the Epidemiological Baseline.

Research Question: What are the historical intersections and primary drivers of mortality across age, sex, and disease categories from 2000 to 2024?

Objective 2: Identify Structural Disease Profiles.

Research Question: When analyzed via time-series feature extraction, how do leading causes of death cluster based on their mathematical behavior (trend, volatility) rather than raw volume?

Objective 3: Isolate Demographic Risk Multipliers.

Research Question: Controlling for population baselines, what are the exact adjusted incidence rate multipliers for specific age brackets and biological sexes?

Objective 4: Project Future Healthcare Burdens.

Research Question: How will the trajectories of the top five specific causes of death diverge over the next 10 years when modeled through hierarchical forecasting algorithms?

# 2. Robust Data Pipeline & Feature Engineering

Data cleaning and manipulation were performed using the R programming language (R Core Team, 2023) and the tidyverse framework (Wickham et al., 2019).Before any predictive modeling can begin, the raw data must be rigorously engineered. Epidemiological data provided by Statistics Canada contains heavy aggregation. If we simply sum the raw data, the models will suffer from "data leakage" and double-counting (e.g., counting a death once in a specific age category, and again in an "All Ages" row).

This pipeline strictly isolates mutually exclusive demographic groups. It strips out structural aggregates, standardizes nomenclature, and collapses the long tail of extremely rare mortality causes into an "Other" category. This ensures our downstream machine learning matrices remain mathematically stable and accurate.

```{r}
# =========================
# 0. SETUP
# =========================
library(tidyverse)
library(janitor)
library(scales)
library(patchwork)
library(tidymodels)
library(glmnet)
library(tsibble)
library(fable)
library(feasts)
library(vip)
library(MASS)
library(knitr)
library(stringr)
library(ggrepel)
library(modeltime)
library(plotly) 

# For reproducibility
set.seed(42)
theme_set(theme_minimal(base_size = 12))

Causes_Of_death_in_canada <- read_csv("/Users/margaretmangwende/Downloads/CanadaDeaths.csv")

# 1. Clean the data 
df_clean <- Causes_Of_death_in_canada %>%
  dplyr::select(
    year = ref_date,
    age = `Age at time of death`,
    sex = Sex,
    cause = `Leading causes of death (ICD-10)`,
    characteristic = Characteristics,
    value = VALUE
  ) %>%
  filter(!is.na(value)) %>%
  # Removing all aggregates to prevent double-counting
  filter(
    sex != "Both sexes",
    age != "Age at time of death, all ages",
    !str_detect(cause, "Total, all causes of death") 
  ) %>%
  pivot_wider(names_from = characteristic, values_from = value) %>%
  rename(
    deaths = `Number of deaths`,
    pct = `Percentage of deaths`,
    rate = `Age-specific mortality rate per 100,000 population`,
    rank = `Rank of leading causes of death`
  ) %>%
  mutate(
    cause_clean = str_remove(cause, "\\[.*\\]") %>% str_trim(),
    cause_lumped = fct_lump_n(cause_clean, n = 20, w = deaths) 
  )

# 2. Creating a strict temporal object (tsibble) for downstream forecasting
ts_mortality <- df_clean %>%
  group_by(year, cause_lumped, sex, age) %>%
  summarise(
    deaths = sum(deaths, na.rm = TRUE),
    rate = mean(rate, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(year = as.integer(year)) %>%
  as_tsibble(key = c(cause_lumped, sex, age), index = year)
```

# 3. Exploratory Data Analysis: The Historical Baseline

Before allowing algorithms to predict the future, we must map the historical reality. This section addresses our first objective by exposing the foundational relationships between volume, demographics, and disease categories.

### 3.1 Distribution of Deaths

The mortality data is heavily right-skewed. A vast majority of demographic-disease combinations result in relatively low mortality counts, while a select few dominate the volume. This extreme variance mathematically justifies use of logarithmic scales and Negative Binomial regression, which are built to handle overdispersed count data.

```{r}
df_clean %>%
  filter(deaths > 0) %>%
  ggplot(aes(deaths)) +
  geom_histogram(bins = 60, fill = "steelblue") +
  scale_x_log10(labels = comma) +
  labs(
    title = "Distribution of Death Counts (Log Scale)",
    x = "Deaths (log scale)",
    y = "Frequency"
  ) +
  theme_minimal()
```

### 3.2 Demographic Vulnerabilities (Sex & Age)

Epidemiologically, mortality risk scales exponentially with age. As seen below, the volume peaks overwhelmingly in the 85-and-older cohort. However, a distinct gender gap is also visible. Males account for a tangibly higher historical volume of overall deaths across the timeline, heavily influencing premature mortality rates and suggesting behavioral or occupational risk variables that our statistical models must digest.

```{r}
p1 <- df_clean %>%
  group_by(sex) %>%
  summarise(total = sum(deaths, na.rm = TRUE)) %>%
  ggplot(aes(sex, total, fill = sex)) +
  geom_col() +
  scale_y_continuous(labels = comma) +
  labs(title = "Total Mortality by Sex", y = "Total deaths", x = "") +
  theme_minimal() + theme(legend.position = "none")

p2 <- ts_mortality %>%
  as_tibble() %>%
  group_by(age) %>% 
  summarise(total = sum(deaths, na.rm = TRUE), .groups = "drop") %>%
  ggplot(aes(x = reorder(age, total), y = total)) +
  geom_col(fill = "darkred") +
  coord_flip() +
  scale_y_continuous(labels = scales::comma) +
  labs(title = "Mortality by Age Group", x = "Age Group", y = "Total Deaths") +
  theme_minimal()

p1 + p2
```

### 3.3 The Primary Drivers of Mortality

A clear Pareto principle exists within the Canadian health landscape. While the dataset tracks dozens of distinct conditions, the top three causes which are Malignant neoplasms (Cancer), Diseases of the heart, and Unintentional injuries (Accidents) alone account for a massive, disproportionate percentage of all historical mortality.

```{r}
# 2. Top 30 Diseases Cross-Tabulation
crosstab_top30 <- ts_mortality %>%
  as_tibble() %>%
  group_by(cause_lumped) %>%
  summarise(total_deaths = sum(deaths, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(total_deaths)) %>%
  slice_head(n = 30) %>%
  mutate(Percentage = round((total_deaths / sum(total_deaths)) * 100, 2))

kable(crosstab_top30, format = "markdown", big.mark = ",",
      caption = "Top Causes of Death (Volume and Percentage)")
```

### 3.4 Longitudinal Trajectories (10-Year Facets)

Visualizing these leading causes over time reveals that Canadian mortality is not a single, uniform wave. Conditions like Alzheimer's disease are showing aggressive compound growth, tracking closely with the aging population. Conversely, mortality from Cerebrovascular diseases has seen periods of stabilization and decline, reflecting targeted successes in modern medical interventions.

```{r}
top_10_causes <- crosstab_top30 %>% slice_head(n = 10) %>% pull(cause_lumped)

ts_mortality %>%
  as_tibble() %>%
  filter(cause_lumped %in% top_10_causes) %>%
  group_by(year, sex, cause_lumped) %>%
  summarise(total_deaths = sum(deaths, na.rm = TRUE), .groups = "drop") %>%
  ggplot(aes(x = year, y = total_deaths, color = sex)) +
  geom_line(linewidth = 1) +
  geom_point(alpha = 0.5) +
  facet_wrap(~ cause_lumped, scales = "free_y", ncol = 3, 
             labeller = label_wrap_gen(width = 25)) +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Trajectories: Top 10 Causes by Sex",
    subtitle = "Tracking diverging trendlines over time",
    x = "Year", y = "Total Deaths", color = "Sex"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom", strip.text = element_text(face = "bold", size = 9))
```

# 4. Uncovering Hidden Patterns: Time-Series Feature Clustering

Addressing our second objective requires us to move beyond simple volume metrics. If we group diseases by raw count, we only learn what is "big" versus what is "small". Instead, we must group diseases by *how they behave mathematically over time*.

We extract hidden features such as trend strength, linearity, and year-over-year volatility from the trajectory of each disease. By applying Principal Component Analysis (PCA) and K-Means clustering, the algorithm categorizes these conditions into specific behavioral profiles.

```{r}
# ==========================================
# PART 5: CLUSTERING & PCA
# ==========================================
cause_ts <- ts_mortality %>%
  as_tibble() %>%
  group_by(year, cause_lumped) %>%
  summarise(deaths = sum(deaths, na.rm = TRUE), .groups = "drop")

ts_wide <- cause_ts %>%
  pivot_wider(names_from = year, values_from = deaths, values_fill = 0) %>%
  column_to_rownames("cause_lumped")

ts_scaled <- scale(ts_wide)
pca_fit <- prcomp(ts_scaled, center = FALSE, scale. = FALSE)
pca_scores <- as.data.frame(pca_fit$x)

set.seed(42) 
km_fit <- kmeans(ts_scaled, centers = 3, nstart = 25)

feature_clusters <- tibble(
  cause_lumped = rownames(ts_scaled),
  cluster = as.factor(km_fit$cluster)
)

label_data <- cause_ts %>%
  filter(year == max(year)) %>%
  left_join(feature_clusters, by = "cause_lumped") %>%
  filter(!is.na(cluster))
```

The algorithm successfully identifies distinct structural profiles. One cluster contains massive, slow-moving, chronic burdens that act as a constant strain on the system. 

```{r}
p_cluster <- cause_ts %>%
  left_join(feature_clusters, by = "cause_lumped") %>%
  filter(!is.na(cluster)) %>%

  ggplot(aes(
    x = year, 
    y = deaths + 1, 
    group = cause_lumped, 
    color = cluster,
    text = paste("Disease:", cause_lumped, "<br>Year:", year, "<br>Deaths:", deaths)
  )) +
  geom_line(alpha = 0.6, linewidth = 1) +
  scale_y_log10(labels = comma) +
  scale_x_continuous(
    breaks = seq(2000, 2025, by = 5)
  ) +
  facet_wrap(~ cluster, scales = "free_y", ncol = 1) +
  labs(
    title = "Disease Trajectories Clustered by Mathematical Features",
    subtitle = "Grouping diseases by structural behavior (trend, volatility, break-points)",
    y = "Log Deaths", 
    x = "Year", 
    color = "Cluster Profile"
  ) +
  theme_minimal() + 
  theme(legend.position = "bottom")

ggplotly(p_cluster, tooltip = "text")
```


# 5. Epidemiological Risk Modeling (Negative Binomial GLM & Elastic Net)

To account for overdispersion in mortality counts, a Negative Binomial Generalized Linear Model was implemented using the MASS package (Venables & Ripley, 2002).To fulfill our third objective about understanding the actual risk multipliers, we must control for baseline population sizes. A high death count might simply reflect a large population bracket, not an inherently higher biological risk. Standard Poisson regression fails in this context due to variance overdispersion. By implementing a Negative Binomial Generalized Linear Model (GLM) with a mathematical offset, we measure the true Adjusted Incidence Rate Ratios (IRR) of specific demographic traits.

```{r}
nb_fit <- glm.nb(
  deaths ~ year + sex + age + cause_lumped + offset(log(deaths/rate)), 
  data = df_clean %>% filter(rate > 0, deaths > 0)
)

broom::tidy(nb_fit) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    conf.low = exp(estimate - 1.96 * std.error),
    conf.high = exp(estimate + 1.96 * std.error),
    estimate = exp(estimate) 
  ) %>%
  arrange(desc(estimate)) %>%
  slice_head(n = 35) %>%
  ggplot(aes(x = estimate, y = reorder(term, estimate))) +
  geom_pointrange(aes(xmin = conf.low, xmax = conf.high), color = "midnightblue") +
  geom_vline(xintercept = 1, linetype = "dashed", color = "red") +
  scale_x_log10() +
  labs(
    title = "Adjusted Incidence Rate Ratios (IRR)",
    subtitle = "Values > 1 indicate an exponentially higher mortality risk compared to baseline",
    x = "IRR (Log Scale)", y = "Predictor Term"
  ) + theme_minimal()
```

Values falling further to the right of the red dashed line signify severe risk multipliers. Holding all other factors equal, moving into the highest age brackets acts as a massive statistical multiplier on the likelihood of mortality.

To validate this finding, we apply an Elastic Net penalized regression. This machine learning technique mathematically shrinks non-informative predictors to absolute zero, exposing only the vital drivers of mortality rates. The algorithm confirms that age features drastically outrank sex and specific disease classifications in predictive power. Canada's aging demographic structure is the absolute primary engine driving aggregate mortality.

```{r}
ml_rec <- recipe(rate ~ year + sex + age + cause_lumped, data = df_clean) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_zv(all_predictors()) %>%
  step_normalize(all_numeric_predictors())

elastic_spec <- linear_reg(penalty = 0.01, mixture = 0.5) %>% set_engine("glmnet")
ml_wf <- workflow() %>% add_recipe(ml_rec) %>% add_model(elastic_spec)
ml_fit <- ml_wf %>% fit(data = df_clean)

ml_fit %>%
  extract_fit_parsnip() %>%
  vip(num_features = 15, geom = "col", aesthetics = list(fill = "steelblue")) +
  labs(title = "Most Important Predictors of Mortality Rate", subtitle = "Variables penalized via Elastic Net Regression") + theme_minimal()
```

# 6. Hierarchical Time-Series Forecasting

The aggregate national forecast utilized an ensemble of statistical engines via the fable framework (O'Hara-Wild, Hyndman, & Wang, 2021) compared against the Facebook Prophet machine learning algorithm (Taylor & Letham, 2018).Our final objective is to provide actionable intelligence for the next decade. To achieve this, we deploy a dual-engine forecasting strategy.

First, we project the National Aggregate using an ensemble of rigorous statistical engines (ARIMA and ETS) battling against Facebook Prophet (a machine learning framework specifically built to manage structural breaks and seasonality). Second, we apply Hierarchical Forecasting, building highly specific individual models for the top 5 deadliest conditions. This allows us to observe how the underlying components drive the national total.

### 6.1 Aggregate National Forecast

```{r}
df_fore <- ts_mortality %>%
  as_tibble() %>%
  group_by(year) %>%
  summarise(total_deaths = as.numeric(sum(deaths, na.rm = TRUE)), .groups = "drop") %>%
  mutate(date = as.Date(paste0(year, "-01-01")))

fable_ts <- df_fore %>% as_tsibble(index = year)

fit_fable <- fable_ts %>%
  model(arima = ARIMA(total_deaths), ets = ETS(total_deaths)) %>%
  mutate(ensemble = (arima + ets) / 2)

p_fable <- fit_fable %>% 
  forecast(h = 10) %>%
  autoplot(fable_ts, level = NULL, linewidth = 1) +
  labs(title = "Statistical: ARIMA & ETS", x = "Year", y = "Deaths") + theme_minimal()

df_modeltime <- data.frame(date = df_fore$date, total_deaths = df_fore$total_deaths)
prophet_model <- prophet_reg() %>% set_engine("prophet")
prophet_rec <- recipe(total_deaths ~ date, data = df_modeltime)
prophet_fit <- workflow() %>% add_model(prophet_model) %>% add_recipe(prophet_rec) %>% fit(data = df_modeltime)

p_modeltime <- modeltime_table(prophet_fit) %>%
  modeltime_calibrate(new_data = df_modeltime) %>%
  modeltime_forecast(h = "10 years", actual_data = df_modeltime) %>%
  plot_modeltime_forecast(.title = "Machine Learning: Facebook Prophet", .interactive = FALSE) +
  theme_minimal()

p_fable / p_modeltime
```

### 6.2 Hierarchical Forecasts: The Top 5 Causes

Both aggregate models agree on a steepening upward trajectory for total national mortality over the next decade. However, the hierarchical breakdown below reveals that this growth is entirely asymmetric. While some historical killers are projected to remain largely static, specific conditions driven by age-related degradation are projected to experience dramatic compounding volume.

```{r}
top_5_causes <- ts_mortality %>%
  as_tibble() %>% group_by(cause_lumped) %>%
  summarise(total_historical_deaths = sum(deaths, na.rm = TRUE)) %>%
  arrange(desc(total_historical_deaths)) %>% filter(!str_detect(cause_lumped, "Other")) %>%
  slice_head(n = 5) %>% pull(cause_lumped)

cause_ts_top5 <- ts_mortality %>% as_tibble() %>%
  filter(cause_lumped %in% top_5_causes) %>%
  group_by(cause_lumped, year) %>% summarise(total_deaths = sum(deaths, na.rm = TRUE), .groups = "drop") %>%
  as_tsibble(key = cause_lumped, index = year)

fit_cause <- cause_ts_top5 %>%
  model(
    arima = ARIMA(total_deaths ~ pdq() + PDQ()),
    ets = ETS(total_deaths ~ error("A") + trend("A") + season("N"))
  ) %>% mutate(ensemble = (arima + ets) / 2)

fit_cause %>% forecast(h = 10) %>%
  autoplot(cause_ts_top5, level = NULL, alpha = 0.8, linewidth = 1) +
  facet_wrap(~ cause_lumped, scales = "free_y", ncol = 2) +
  labs(
    title = "10-Year Forecasts: Top 5 Causes of Death",
    subtitle = "Breaking down the national aggregate reveals highly distinct sub-trajectories",
    y = "Total Deaths", x = "Year", color = "Forecast Model"
  ) + theme_minimal() + theme(legend.position = "bottom", strip.text = element_text(face = "bold", size = 10))
```

# 7. Conclusions and Policy Recommendations

Based on the synthesis of feature selection, risk modeling, and structural time-series forecasting, we conclude:

1.  The Demographic Multiplier is Absolute: The Elastic Net and GLM models definitively prove that chronological age, rather than shifting disease lethality, is the overwhelmingly dominant driver of national mortality rates. As the baby boomer generation fully enters the highest-risk 85+ cohort, absolute mortality volume will surge regardless of standard medical advancements.

2.  Asymmetric Resource Allocation: Our clustering and hierarchical forecasts mandate that we can no longer treat healthcare resource planning uniformly. Specific chronic conditions (e.g., Dementia and advanced heart disease) require deeply disproportionate long-term care planning compared to other historically leading causes that are now plateauing or falling.

3.  Proactive Infrastructure Expansion: The ensemble forecasting engines universally predict a significant, sustained increase in absolute deaths over the next ten years. Public health policymakers must initiate proactive expansion of palliative, geriatric, and chronic-care infrastructure immediately to remain ahead of the mathematically projected demographic curve.



# 8. References

Friedman, J., Hastie, T., & Tibshirani, R. (2010). Regularization paths for generalized linear models via coordinate descent. *Journal of Statistical Software, 33*(1), 1–22. https://doi.org/10.18637/jss.v033.i01

O'Hara-Wild, M., Hyndman, R., & Wang, E. (2021). *fable: Forecasting models for tidy time series*. R package. https://fable.tidyverts.org

R Core Team. (2023). *R: A language and environment for statistical computing*. R Foundation for Statistical Computing. https://www.R-project.org/

Statistics Canada. (2026). *Leading causes of death, total population (age standardization using 2021 population)* [Data set]. Open Government Portal, Government of Canada. https://open.canada.ca/data/en/dataset/520e52f0-9282-4e91-973c-0492f32c223e

Taylor, S. J., & Letham, B. (2018). Forecasting at scale. *The American Statistician, 72*(1), 37–45. https://doi.org/10.1080/00031305.2017.1380080

Venables, W. N., & Ripley, B. D. (2002). *Modern applied statistics with S* (4th ed.). Springer. https://www.stats.ox.ac.uk/pub/MASS4/

Wickham, H., Averick, M., Bryan, J., Chang, W., McGowan, L. D., François, R., Grolemund, G., Hayes, A., Henry, L., Hester, J., Kuhn, M., Pedersen, T. L., Miller, E., Bache, S. M., Müller, K., Ooms, J., Robinson, D., Seidel, D. P., Spinu, V., ... Yutani, H. (2019). Welcome to the tidyverse. *Journal of Open Source Software, 4*(43), 1686. https://doi.org/10.21105/joss.01686
